---
title: "AIDS progression"
output: rmarkdown::html_vignette
# output: rmarkdown::pdf_document
rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    code_folding: show
bibliography: hdbayes.bib  
vignette: >
  %\VignetteIndexEntry{Logistic_regression_HIV}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Here we will analyse the data in Section 4.2 of @Chen1999 using a logistic regression 
model. 
The analysis consists of two AIDS clinical trials called ACTG036 and ACTG019,
the data for which can be accessed by calling `data(actg036)` and
`data(actg019)`, respectively (see below).

The historical data will come from the ACTG019 (@Volberding1990) study, which
was a double-blind placebo-controlled clinical trial comparing zidovudine (AZT)
with a placebo in people with CD4 cell counts less than 500.
The sample size of complete observations for this study was `n0` = 822.
The response variable (`y0`) for these data is binary, with 1 indicating death,
development of AIDS or ARC and 0 otherwise.
We will use the following covariates:  `CD4` cell count, `age`, `treatment` and
`race`.
To facilitate computation, we will center and scale the continuous covariates (`age` and `CD4` count).
In general, we recommend this centering procedure in order to keep coefficients on roughly the same scale and thus avoid difficult posterior geometries for the Stan dynamic Hamiltonian Monte Carlo (dHMC) to explore.

We will use the methods in **hdbayes** to construct informative priors using the
ACTG019 data as historical data in order to analyse the data from ACTG036 study
(@Merigan1991), for which we will use the same four covariates.
For the ACTG036 study the sample size was `n`= 183.

Let's take a look at the data and take the opportunity to centre the continuous
covariates (`CD4` and `age`):

```{r setup, cache = TRUE}
library(parallel)
library(posterior)
library(ggplot2)
library(ggthemes)
library(tibble)
library(wacolors)
library(tidyverse)
library(hdbayes)
```

```{r setup_data, cache = TRUE}

data("actg019")
data("actg036")

summary(actg019)
summary(actg036)

actg019$age <- scale(actg019$age)
actg019$cd4 <- scale(actg019$cd4)

actg036$age <- scale(actg036$age)
actg036$cd4 <- scale(actg036$cd4)
```

To establish a baseline, we will estimate the parameters via maximum likelihood,
using `glm()` from **stats**:

```{r mle, cache = TRUE}
formula <- outcome ~  age + race + treatment + cd4
p <- length(attr(terms(formula), "term.labels")) # number of predictors
family  <- binomial('logit')

fit.mle.cur  <- glm(formula, family, actg036)
fit.mle.hist <- glm(formula, family, actg019)

summary(fit.mle.hist)
summary(fit.mle.cur)

confint(fit.mle.hist)
confint(fit.mle.cur)

the.data <-  list(actg036,
                  actg019)

```
from which we can see quite some discrepancy in the coefficient estimates.
In particular, there is substantial uncertainty in the estimates of the treatment
effect, as evidenced by a wide 95% confidence interval.
We would thus like to incorporate information from the ACTG019 trial into a prior
distribution for the regression coefficients.
We will use the plethora of methods available in **hdbayes** to construct informative priors using the available historical data.

### Bayesian Hierarchical model (BHM)

Our first approach will be to fit a Bayesian hierarchical model (BHM), where we model
the coefficients for historical and current data jointly through a hierarchical (multilevel) structure: the coefficients for each study are drawn from the same multivariate normal distribution.
The hyperparameters of this distribution are also assigned (inverse Wishart and inverse Gamma) hyperpriors.

In summary, we employ the model:

\begin{align*}
y_i | x_i, \beta &\sim \text{Bernoulli}\left( \text{logit}^{-1}(x_i'\beta) \right),\\
y_{0i} | x_{0i}, \beta_0 &\sim \text{Bernoulli}\left( \text{logit}^{-1}(x_{0i}'\beta_0) \right), \\
\beta, \beta_0 &\sim N_p(\mu, \Sigma), \\
\mu &\sim N_p(\mu_0, \Sigma_0), \\
\Sigma &\sim \text{IW}_p(\nu_0, \Psi_0).
\end{align*}

where by default we set

* $\mu_0 = 0;$
* $\Sigma_0 = I_p;$
* $\nu_0 = p + 10;$
* $\Psi_0 = I_p,$

where $p$ is the number of predictors (including the intercept if applicable).

To fit this model, let us first set up the computational specs, which will also be
used in the subsequent analyses.

```{r compsetup, cache = TRUE}
ncores <- 4 # max(1, parallel::detectCores() - 2)
nchains <- ncores
warmup  <- 1000
total.samples <- 10000   ## number of samples post warmup
samples <- ceiling( total.samples / ncores)  ## outputs approx total.samples samples
base.pars <- c("(Intercept)", "age", "race", "treatment", "cd4")

get_summaries <- function(fit, pars.interest,
                          digits = 2) {
  ## A little function to pull out the summaries in a convenient form
  out <- subset(posterior::summarise_draws(fit,
                                           .num_args = list(
                                             sigfig = digits,
                                             notation = "dec"
                                           )),
                variable %in% pars.interest)
  
  return(out)
}

```

Now, a simple call to `glm.bhm()` will fit the model:
```{r bhm, cache = TRUE}
time.bhm <- system.time(
  fit.bhm <- glm.bhm(
    formula,
    family,
    data.list = the.data,
    meta.mean.mean = 0,
    meta.mean.sd = 10,
    meta.sd.mean = 0,
    meta.sd.sd = .5,
    parallel_chains = 4,
    iter_warmup = warmup,
    iter_sampling = samples,
    refresh = 0
  )
)
get_summaries(fit.bhm, pars.interest = base.pars)
```
As we can see, all MCMC (Stan) diagnostics (Rhat, divergences, etc) look good so we can continue.

### Commensurate prior

Next, we consider the commensurate prior of @Hobbs2012.
This model has the following structure:
\begin{align*}
y_i | x_i, \beta &\sim \text{Bernoulli}\left( \text{logit}^{-1}(x_i'\beta) \right) ,\\
y_{0i} | x_{0i}, \beta_0 &\sim \text{Bernoulli}\left( \text{logit}^{-1}(x_{0i}'\beta_0) \right), \\
\beta_0 &\sim N_p(\mu_0, \Sigma_0), \\
\beta_j &\sim N_1\left( \beta_{0j}, \tau_j^{-1} \right), j = 1, \ldots, p,
\end{align*}
where the $\tau_j$'s are precisions elicited by the user.
The defaults in **hdbayes** are

* $\mu_0 = 0$
* $\Sigma_0 = 100 \times I_p$

Again, a simple call to the appropriate function
```{r commensurate, cache=TRUE}
time.commensurate <- system.time(
  fit.commensurate <- glm.commensurate(
    formula,
    family,
    tau = rep(5, p + 1),
    data.list = the.data,
    parallel_chains = 4,
    iter_warmup = warmup,
    iter_sampling = samples,
    refresh = 0
  )
)

get_summaries(fit.commensurate, pars.interest = base.pars)
```
will fit the model.
And again the diagnostics are fine, so we may proceed with our exploration of
informative priors.

### Robust Meta-Analytic Predictive (RMAP) Prior

The Robust MAP prior (@Schmidli2014) is a generalization of the Bayesian Hierarchical Model (BHM), and takes the form
\begin{align*}
y_i | x_i, \beta &\sim \text{Bernoulli}\left( \text{logit}^{-1}(x_i'\beta) \right), \\
y_{0i} | x_{0i}, \beta_0 &\sim \text{Bernoulli}\left( \text{logit}^{-1}(x_{0i}'\beta_0) \right), \\
\beta_0 &\sim N_p(\mu, \Sigma), \\
\beta   &\sim w \times N_p(\mu, \Sigma) + (1 - w) N_p(\mu_v, \Sigma_v), \\
\mu &\sim N_p(\mu_0, \Sigma_0), \\
\Sigma &\sim \text{IW}_p(\nu_0, \Psi_0),
\end{align*}

where $w \in (0,1)$ controls for the level of borrowing of the historical data.
Note that when $w = 1$, the robust MAP prior effectively becomes the BHM.
The defaults are the same as in the BHM and the default value for $w$ is 0.1.
See the [sensitivity analysis vignette](SensitivityAnalysis.html) for insight into how
to set $w$.
The RMAP can be called with `glm.robustmap()`:
```{r robustmap, cache = TRUE}
rmap.t1 <- system.time(
  fit.bhm.hist <- glm.rmap.bhm(
    formula,
    family,
    hist.data.list = list(actg019),
    meta.mean.mean = 0,
    meta.mean.sd = 10,
    meta.sd.mean = 0,
    meta.sd.sd = .5,
    parallel_chains = 4,
    iter_warmup = warmup,
    iter_sampling = samples,
    thin = 10,
    refresh = 0
  )
)
samples_bhm <- fit.bhm.hist$beta_pred

rmap.t2 <- system.time(res_approx <- glm.rmap.bhm.approx(
  samples.bhm = samples_bhm,
  G = 1:9,
  verbose = FALSE
))

rmap.t3 <- system.time(
  fit.rmap.a <- glm.rmap(
    formula,
    family,
    curr.data = actg036,
    probs = res_approx$probs,
    means = res_approx$means,
    covs = res_approx$covs,
    w = 0.1,
    parallel_chains = 4,
    iter_warmup = warmup,
    iter_sampling = samples,
    refresh = 0
  )
)

rmap.t4 <- system.time(
  fit.rmap.b <- glm.rmap(
    formula,
    family,
    curr.data = actg036,
    probs = res_approx$probs,
    means = res_approx$means,
    covs = res_approx$covs,
    w = 0.5,
    parallel_chains = 4,
    iter_warmup = warmup,
    iter_sampling = samples,
    refresh = 0
  )
)

rmap.t5 <- system.time(
  fit.rmap.c <- glm.rmap(
    formula,
    family,
    curr.data = actg036,
    probs = res_approx$probs,
    means = res_approx$means,
    covs = res_approx$covs,
    w = 0.9,
    parallel_chains = 4,
    iter_warmup = warmup,
    iter_sampling = samples,
    refresh = 0
  )
  
)

time.rmap <- rmap.t1 + rmap.t2 + rmap.t4

get_summaries(fit.rmap.a,
              pars.interest = base.pars)
get_summaries(fit.rmap.b,
              pars.interest = base.pars)
get_summaries(fit.rmap.c,
              pars.interest = base.pars)
```
The diagnostics look good, so we will proceed.

### Power priors

In this section we will explore the power prior (PP, @Ibrahim2000) and its variations: the normalised  power prior (NPP, @Duan2006, @Neuenschwander2009) and the normalised asymptotic power prior (NAPP, @Ibrahim2015).
The Power Prior takes the form
\begin{align*}
y_i | x_i, \beta &\sim \text{Bernoulli}\left( \text{logit}^{-1}(x_i'\beta) \right), \\
y_{0i} | x_{0i}, \beta &\sim \text{Bernoulli}\left( \text{logit}^{-1}(x_{0i}'\beta) \right), \\
\pi(\beta | a_0) &\propto L(\beta | y_0)^{a_0} \pi_0(\beta),
\end{align*}
where $L(\beta | y_0)$ is the likelihood of the GLM based on the historical data,
$a_0 \in (0,1)$ is a fixed hyperaparameter controlling the effective
sample size contributed by the data (e.g., $a_0 = 1$ borrows the whole sample size),
and $\pi_0(\beta)$ is an "initial prior" on $\beta$.

The default in **hdbayes** is a (noninformative) normal prior on $\beta$:
$$
\beta \sim N_p(0, 100 \times I_p).
$$

The question that then arises is how to choose $a_0$.
While no definitive answer can be given without context of the specific data and model
under analysis, we find it reasonable to set $a_0 = a^\star = \frac{1}{2}\frac{n}{n_0}$.
This way, when the historical and current data sets are of the same size, we set the borrowing factor to $a_0 = 1/2$, reflecting no particular desire to either borrow or not borrow information.
Let us do just that
```{r astar, cache = TRUE}
n0 <- nrow(actg019)
n <- nrow(actg036)
a0.star <- (n/n0) * 1/2
```
and then proceed to call `glm.pp()` twice (once for $a_0 = 1/2$ and once for $a_0 = a^\star$):

```{r pp, cache = TRUE}
time.pp <- system.time(
  fit.pp <- glm.pp(
    formula,
    family,
    data.list = the.data,
    a0 = 0.5,
    parallel_chains = 4,
    iter_warmup = warmup,
    iter_sampling = samples,
    refresh = 0
  )
)

time.pp.star <- system.time(
  fit.pp.star <- glm.pp(
    formula,
    family,
    data.list = the.data,
    a0 = a0.star,
    parallel_chains = 4,
    iter_warmup = warmup,
    iter_sampling = samples,
    refresh = 0
  )
)

get_summaries(fit.pp,
              pars.interest = base.pars)
get_summaries(fit.pp.star,
              pars.interest = base.pars)
```
from which we can see that (i) all of the diagnostics are fine and (ii) the choice 
of $a_0$ does seem to matter (look at the estimates for the `treatment` coefficient).

#### Normalised power prior (NPP)

Now we are prepared to start our normalised power prior analysis.
The NPP treats the hyperparameter $a_0$ as random, allowing the data to decide
what is the best value.
For most models, this requires estimating the normalising constant
$Z(a_0) = \int L(\beta | y_0)^{a_0} \pi_0(\beta) d\beta$.

The NPP may be summarized as
\begin{align*}
y_i | x_i, \beta &\sim \text{Bernoulli}\left( \text{logit}^{-1}(x_i'\beta) \right), \\
y_{0i} | x_{0i}, \beta &\sim \text{Bernoulli}\left( \text{logit}^{-1}(x_{0i}'\beta) \right), \\
\pi(\beta | a_0) &\propto \frac{1}{Z(a_0)} L(\beta | y_0)^{a_0} \pi_0(\beta), \\
\pi(a_0)         &\propto a_0^{\alpha_0 - 1} (1 - a_0)^{\gamma_0 - 1}.
\end{align*}

The defaults in **hdbayes** are

* $\pi_0(\beta) \propto N(\beta | 0, 100 \times I_p)$;
* $\alpha_0 = 1$;
* $\gamma_0 = 1$,

when $\alpha_0 = 1$ and $\gamma_0 = 1$, the prior on $a_0$ is a $U(0,1)$
prior.

We thus might decide to let $a_0$ vary and be estimated from data.
This naturally necessitates placing a prior on $a_0$.
We will now construct a Beta prior for $a_0$ centred on $a^\star = \frac{1}{2}\frac{n}{n_0}$ with coefficient of variation (cv) equal to $1$:

```{r a0prior, cache = TRUE, fig.align='center', fig.height=4, fig.width=8, fig.fullwidth=TRUE}
devtools::source_url("https://raw.githubusercontent.com/maxbiostat/logPoolR/main/R/beta_elicitator.r")
beta.pars <- elicit_beta_mean_cv(m0 = a0.star, cv = 1)
curve(dbeta(x, shape1 = beta.pars$a, shape2 = beta.pars$b), 
      lwd = 3,
      main = "Prior on a0",
      ylab = "Density",
      xlab = expression(a[0]))
abline(v = a0.star, lwd = 2, lty = 3)
legend(x = "topright",
       legend = c(expression(pi(a[0])),
                  expression(a^'*')),
       lwd = 2, lty = c(1, 3),
       bty = 'n')
```

To conduct an NPP analysis we first need to estimate the normalising constant
$c(a_0)$ for a grid of values of $a_0$.
In **hdbayes**, there is one function to estimate the normalising constant across
a grid of values for $a_0$ and another to obtain posterior samples of the
normalized power prior.

```{r normconst, cache = TRUE}
a0       <- seq(0, 1, length.out = 11)
a0.lognc <- list()
a0.lognc.hdbayes <- data.frame(a0 = a0)

glm_lp <- hdbayes:::glm_lp
get_lp2mean <- hdbayes:::get_lp2mean
normal_glm_lp <- hdbayes:::normal_glm_lp
bernoulli_glm_lp <- hdbayes:::bernoulli_glm_lp
poisson_glm_lp <- hdbayes:::poisson_glm_lp
gamma_glm_lp <- hdbayes:::gamma_glm_lp
invgauss_glm_lp <- hdbayes:::invgauss_glm_lp


## call created function
for (i in 2:length(the.data)) {
  histdata = the.data[[i]]
  ## wrapper to obtain log normalizing constant in parallel package
  logncfun <- function(a0, ...) {
    glm.npp.lognc(
      formula = formula,
      family = family,
      a0 = a0,
      histdata = histdata,
      ...
    )
  }
  cl <- parallel::makeCluster(10)
  parallel::clusterSetRNGStream(cl, 123)
  parallel::clusterExport(
    cl,
    varlist = c(
      'formula',
      'family',
      'histdata',
      'glm.npp.lognc',
      'glm_lp',
      'get_lp2mean',
      'normal_glm_lp',
      'bernoulli_glm_lp',
      'poisson_glm_lp',
      'gamma_glm_lp',
      'invgauss_glm_lp'
    )
  )
  time.npp.1 <- system.time(
    a0.lognc[[i - 1]] <- parallel::parLapply(
      cl = cl,
      X = a0,
      fun = logncfun,
      iter_warmup = warmup,
      iter_sampling = samples,
      chains = 1,
      refresh = 0
    )
  )
  
  stopCluster(cl)
  a0.lognc[[i - 1]] <- data.frame(do.call(rbind, a0.lognc[[i - 1]]))
  a0.lognc.hdbayes <- cbind(a0.lognc.hdbayes, a0.lognc[[i - 1]]$lognc)
  colnames(a0.lognc.hdbayes)[i] <- paste0("lognc_hist", i - 1)
}

a0.lognc <- a0.lognc.hdbayes$a0
lognc    <- as.matrix(a0.lognc.hdbayes[, -1, drop = F])
```
We will now fit the NPP with both a uniform (Beta(1, 1)) and the informative prior 
on $a_0$ devised above using the dictionary of $a_0$ and $c(a_0)$ we just created:

```{r nppfits, cache = TRUE}
time.npp.2 <- system.time(
  fit.npp_unif <- glm.npp(
    formula,
    family,
    data.list = the.data,
    a0.lognc = a0.lognc,
    lognc = lognc,
    parallel_chains = 4,
    iter_warmup = warmup,
    iter_sampling = samples,
    refresh = 0
  )
)

devtools::source_url("https://raw.githubusercontent.com/maxbiostat/logPoolR/main/R/beta_elicitator.r")
beta.pars <- elicit_beta_mean_cv(m0 = a0.star, cv = 1)


time.npp.2.star <- system.time(
  fit.npp.star <- glm.npp(
    formula,
    family,
    data.list = the.data,
    a0.lognc = a0.lognc,
    lognc = lognc,
    a0.shape1 = beta.pars$a, a0.shape2 = beta.pars$b,
    parallel_chains = 4,
    iter_warmup = warmup,
    iter_sampling = samples,
    refresh = 0
  )
)


get_summaries(fit.npp_unif,
              pars.interest = base.pars)
get_summaries(fit.npp.star,
              pars.interest = base.pars)
```
With all diagnostics failing to detect problems we move on.

#### Normalized asymptotic power prior (NAPP)

The Normalized asymptotic power prior (NAPP) uses a large sample theory argument to formulate a normal approximation to the power prior, i.e., the prior is given by
$$
\beta | a_0 \sim N(\hat{\beta}_0, a_0^{-1} [I_n(\beta)]^{-1}),
$$
where $\hat{\beta}_0$ is the maximum likelihood estimate (MLE) of $\beta$
based on the historical data and $I_n(\beta)$ is the associated information
matrix (negative Hessian).

In this case, the normalising constant is known, so we do not need to estimate it before sampling.

The NAPP can be fitted with a call to `glm.napp()`:
```{r nappfits, cache = TRUE}
time.napp <- system.time(
  fit.napp_unif <- glm.napp(
    formula,
    family,
    data.list = the.data,
    parallel_chains = 4,
    iter_warmup = warmup,
    iter_sampling = samples,
    refresh = 0
  )
)

time.napp.star <- system.time(
  fit.napp.star <- glm.napp(
    formula,
    family,
    data.list = the.data,
    a0.shape1 = beta.pars$a, a0.shape2 = beta.pars$b,
    parallel_chains = 4,
    iter_warmup = warmup,
    iter_sampling = samples,
    refresh = 0
  )
)

get_summaries(fit.napp_unif,
              pars.interest = base.pars)
get_summaries(fit.napp.star,
              pars.interest = base.pars)
```

###  Comparison of methods
After all this work, we can now finally compare the point estimate (MLE / posterior mean) and uncertainty (SE / posterior standard deviation) of all the methods considered here.

```{r comparison}

fit.list <- list('BHM' = fit.bhm,
                 'Commensurate' = fit.commensurate,
                 'RMAP_w=0.1' = fit.rmap.a,
                 'RMAP_w=0.5' = fit.rmap.b,
                 'RMAP_w=0.9' = fit.rmap.c,
                 'NAPP_unif' = fit.napp_unif,
                 'NAPP' = fit.napp.star,
                 'NPP_unif' = fit.npp_unif,
                 'NPP' = fit.npp.star,
                 'PP_half' = fit.pp,
                 'PP' = fit.pp.star)

```
As we can see, there seems to be quite some variation in estimates across methods.
To aid our understanding, we will now visualise the posterior distributions for
each method:

```{r prepfigs, echo=FALSE, cache=TRUE}
get_quants <- function(x, name){
  pars <- names(coef(fit.mle.cur))
  summs <- get_summaries(x, pars.interest = pars)
  out <- tibble::tibble(
    parameter = pars,
    mean = as.numeric(unlist(summs[, "mean"])),
    sd = as.numeric(unlist(summs[, "sd"])),
    lwr = as.numeric(unlist(summs[, "q5"])),
    upr = as.numeric(unlist(summs[, "q95"]))
  )
  out$model <- name
  return(out)
} 


model.names <- names(fit.list)

results.list <- lapply(seq_along(fit.list), function(i){
  get_quants(x = fit.list[[i]], name = model.names[i])
})

results <- do.call(rbind, results.list)

mle.estimates.past <- data.frame(
  parameter = names(coef(fit.mle.hist)),
  value = coef(fit.mle.hist),
  MLE = "Historical"
)
mle.estimates.curr <- data.frame(
  parameter = names(coef(fit.mle.cur)),
  value = coef(fit.mle.cur),
  MLE = "Current"
)
```
```{r posteriorplots, echo=FALSE, cache=TRUE,  fig.align='left', fig.height=6, fig.width=9, fig.fullwidth=TRUE}

results$model <- as.factor(results$model)
LL <- unique(results$model)

# run this: wacolors::pal_vector("rainier", n = length(fit.list))

Colours <- setNames( c("#465177", "#CDC24D", "#E1B11F", "#AF6522", "#613E30",
                       "#284739", "#497734", "#85A95C", "#9CB9BF", "#B19EEA",
                       "#DF3383"),
                     LL)


mle.estimates <- rbind(mle.estimates.past,
                       mle.estimates.curr)

not.plot <- c("NPP_unif", "NAPP_unif",
              "PP_half",
              "RMAP_w=0.5", "RMAP_w=0.9")

ggplot() + 
  geom_pointrange(data = subset(results, !model %in% not.plot ),
                  mapping = aes(x = model,
                                y = mean,
                                ymin = lwr, ymax = upr,
                                colour = model)) +
  scale_color_manual(values = Colours) + 
  geom_hline(data = mle.estimates,
             aes(yintercept = value, linetype = MLE)) +
  scale_linetype_manual(values = c("dotted", "dashed")) + 
  scale_y_continuous("") +
  scale_x_discrete("") +
  facet_wrap(parameter~., scales = "free_y") +
  ggtitle("AIDS example - logistic regression") +
  theme_bw(base_size = 20) +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank())


```
where we have omitted the NPP and NAPP with uniform priors and also the power
prior with $a_0 = 1/2$ for clarity.
The first observation to make is that the BHM and robust MAP shrink the most 
towards the historical MLE for the `treatment` effect, but the opposite behaviour
occurs for the coefficient for `race`.
Let's take a look at the coefficients under the rMAP with different values for $w$ ($w=0.1$, $w=0.5$ and $w=0.9$).

```{r rmap_plots, echo=FALSE, cache=TRUE,  fig.align='left', fig.height=6, fig.width=9, fig.fullwidth=TRUE}

rmapstuff <- c("RMAP_w=0.1",
               "RMAP_w=0.5",
               "RMAP_w=0.9")

ggplot() + 
  geom_pointrange(
    data = subset(results, model %in% c(rmapstuff, "BHM") ),
    mapping = aes(x = model,
                  y = mean,
                  ymin = lwr, ymax = upr,
                  colour = model)) +
  scale_y_continuous("") +
  scale_x_discrete("") +
  scale_color_manual(values = Colours) + 
  geom_hline(data = mle.estimates,
             aes(yintercept = value, linetype = MLE)) +
  scale_linetype_manual(values = c("dotted", "dashed")) + 
  facet_wrap(parameter~., scales = "free_y") +
  ggtitle("AIDS example - robust meta-analytic priors") +
  theme_bw(base_size = 20) +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank())
```


We can now plot the posterior summaries for the coefficients under the NPP and
NAPP with uniform and informative priors on $a_0$ in order to visualise the effect
these different prior choices have on the estimated coefficients.

```{r nxpp_plots, echo=FALSE, cache=TRUE,  fig.align='left', fig.height=6, fig.width=9, fig.fullwidth=TRUE}

normalised <- c("NPP", "NPP_unif",
                "NAPP", "NAPP_unif")
ggplot() + 
  geom_pointrange(
    data = subset(results, model %in% normalised),
    mapping = aes(x = model,
                  y = mean,
                  ymin = lwr, ymax = upr,
                  colour = model)) +
  scale_y_continuous("") +
  scale_x_discrete("") +
  scale_color_manual(values = Colours) + 
  geom_hline(data = mle.estimates,
             aes(yintercept = value, linetype = MLE)) +
  scale_linetype_manual(values = c("dotted", "dashed")) + 
  facet_wrap(parameter~., scales = "free_y") +
  ggtitle("AIDS example - normalised power priors") +
  theme_bw(base_size = 20) +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank())
```
Finally, we will take a look at the posterior distribution of $a_0$ for the NPP
and NAPP models under uniform and informative priors:

```{r a0post_plots, echo=FALSE, cache=TRUE,  fig.align='left', fig.height=6, fig.width=9, fig.fullwidth=TRUE}
a0.posteriors <- do.call(rbind, list(
  tibble(a0 = fit.napp.star$a0_hist_1,
         model = "NPP"),
  tibble(a0 = fit.npp_unif$a0_hist_1,
         model = "NPP_unif"),
  tibble(a0 = fit.napp.star$a0_hist_1,
         model = "NAPP"),
  tibble(a0 = fit.napp_unif$a0_hist_1,
         model = "NAPP_unif")
)) 

ggplot() + 
  geom_density(
    data = a0.posteriors,
    mapping = aes(x = a0, colour = model, fill = model),
    alpha = 0.4) +
  scale_color_manual(values = Colours) +
  scale_fill_manual(values = Colours) +
  stat_function(fun = function(x) dbeta(x, beta.pars$a,
                                        beta.pars$b),
                geom = "line", colour = "black",
                linetype = "longdash") + 
  geom_vline(xintercept = a0.star, linetype = "dotted") + 
  scale_y_continuous("", expand = c(0, 0)) +
  scale_x_continuous(expression(a[0]), expand = c(0, 0))+ 
  theme_bw(base_size = 20)
```

where the vertical dotted line marks $a^\star = \frac{1}{2}\frac{n}{n_0}$ and the 
dashed curve depicts the informative beta prior.
As we can see, the prior does make a big difference with regards to the posterior
for $a_0$.
This is not unexpected, since $a_0$ is a hierarchical parameter and learning it 
from data is not a simple task.

Now a comparison of running times

```{r timecomparison}
time.bhm
time.commensurate
time.rmap
time.pp.star
time.npp.1 + time.npp.2
time.npp.1 + time.npp.2.star
time.napp
time.napp.star
```

# References {-}

<div id="refs"></div>

# Computing environment

```{r env}
sessionInfo()
```
