---
title: "Evaluating the sensitivity of informative priors based on historical data"
output: rmarkdown::html_vignette
# output: rmarkdown::pdf_document
bibliography: hdbayes.bib  
rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    code_folding: show
vignette: >
  %\VignetteIndexEntry{SensitivityAnalysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  \VignetteDepends{posterior, ggplot2, ggthemes, tidyverse, scales}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```
The power prior (PP) and the robust meta-analytic prior (RMAP) are two very 
common choices for constructing informative priors by borrowing information from
historical data.
However, these modelling devices depend on tuning parameters that must be chosen
in advance, and it is not clear which values should be picked.
In this vignette we will use the data set in the
[AIDS progression](Logistic_regression_HIV.html) to explore the choice of tuning
parameter in a logistic regression setting.
Please refer to that document for model, data and research question details.

```{r prep, echo=FALSE}
library(hdbayes)
library(posterior)
library(ggplot2)
library(ggthemes)
library(tidyverse)
library(scales)

get_summaries <- function(fit, pars.interest,
                          digits = 2) {
  ## A little function to pull out the summaries in a convenient form
  out <- subset(posterior::summarise_draws(fit,
                                           .num_args = list(
                                             sigfig = digits,
                                             notation = "dec"
                                           )),
                variable %in% pars.interest)
  
  return(out)
}

```

Let' setup
```{r setup_data, cache=TRUE}

data("actg019")
data("actg036")

actg019$age <- scale(actg019$age)
actg019$cd4 <- scale(actg019$cd4)

actg036$age <- scale(actg036$age)
actg036$cd4 <- scale(actg036$cd4)

ncores <- 4 # max(1, parallel::detectCores() - 2)
nchains <- ncores
warmup  <- 1000
total.samples <- 10000   ## number of samples post warmup
samples <- ceiling(total.samples / ncores)  ## outputs approx total.samples samples
```

And obtain some maximum likelihood estimates (MLE):

```{r mle, cache = TRUE}
formula <- outcome ~  age + race + treatment + cd4
p <- length(attr(terms(formula), "term.labels")) # number of predictors
family  <- binomial('logit')

fit.mle.cur  <- glm(formula, family, actg036)
fit.mle.hist <- glm(formula, family, actg019)

mle.estimates.past <- data.frame(
  parameter = names(coef(fit.mle.hist)),
  value = coef(fit.mle.hist),
  MLE = "Historical"
)
mle.estimates.curr <- data.frame(
  parameter = names(coef(fit.mle.cur)),
  value = coef(fit.mle.cur),
  MLE = "Current"
)

mle.estimates <- rbind(mle.estimates.past,
                       mle.estimates.curr)

confint(fit.mle.hist)
confint(fit.mle.cur)

base.pars <- c("(Intercept)", "age", "race", "treatment", "cd4")

the.data <-  list(actg036,
                  actg019)
```

# Robust Meta-analytic predictive prior (RMAP)

The RMAP (@Schmidli2014) is a (linear) mixture between the Bayesian hierarchical model (BHM) and 
a vague prior.
Conceptually, if $\theta$ is the parameter of interest, we have
\begin{equation}
\pi_{\text{RMAP}}(\theta) = w\pi_{\text{BHM}}(\theta) + (1-w)\pi_{\text{Vague}}(\theta),
\end{equation}
where $w \in (0,1)$ is a mixture weight that controls for the level of borrowing of the historical data.
Note that when $w = 1$, the robust MAP prior should collapse to the BHM.
The defaults are the same as in the BHM and the default value for $w$ is 0.1.
We will now fit the RMAP for a few values of $w$ and see what effect (if any)
this has on the regression coefficients (hence $\theta = \boldsymbol{\beta}$ here).
First, let us wrap the function to fit the RMAP,  `glm.robustmap()`, into a
convenient function that takes a given $w$ and returns the output:
```{r mapsetup, cache=TRUE}

fit.bhm.hist <- glm.rmap.bhm(
  formula,
  family,
  hist.data.list = list(actg019),
  meta.mean.mean = 0,
  meta.mean.sd = 10,
  meta.sd.mean = 0,
  meta.sd.sd = .5,
  parallel_chains = 4,
  iter_warmup = warmup,
  iter_sampling = samples,
  thin = 10,
  refresh = 0
)

samples_bhm <- fit.bhm.hist$beta_pred

res_approx <- glm.rmap.bhm.approx(
  samples.bhm = samples_bhm,
  G = 1:9,
  verbose = FALSE
)
```

Now, we will do this for a small set of weights, just to see what is going on:
```{r regularRMAP, cache=TRUE}

rmap_fit_and_report <- function(weight){
  cat("Doing w=", weight, "\n")
  fit.robustmap <- glm.rmap(
    formula,
    family,
    curr.data = actg036,
    probs = res_approx$probs,
    means = res_approx$means,
    covs = res_approx$covs,
    w = weight,
    parallel_chains = 4,
    iter_warmup = warmup,
    iter_sampling = samples,
    show_messages = FALSE,
    show_exceptions = FALSE,
    refresh = 0
  )
  return(fit.robustmap)
}

ws.regular <- seq(0.1, 0.9, by = 0.1) 

system.time(
  all.fits.regular <- lapply(ws.regular, rmap_fit_and_report)
)

all.summaries.regular <- lapply(1:length(all.fits.regular),  function(i) {
  out <- tibble(get_summaries(fit = all.fits.regular[[i]], base.pars),
                name = paste0("w=", ws.regular[i]))
  return(out)
})

results.regular <- do.call(rbind, all.summaries.regular) 
results.regular$w <- as.numeric(gsub("w=", "", results.regular$name))
results.regular <- results.regular %>% rename(lwr = q5)
results.regular <- results.regular %>% rename(upr = q95)
results.regular <- results.regular %>% rename(parameter = variable)
```
Next, we will produce a little plot to help us visualise things:

```{r RMAPplot_regular, echo=FALSE, cache=TRUE,  fig.align='left', fig.height=6, fig.width=9, fig.fullwidth=TRUE}

ggplot(data = results.regular,
       aes(x = w, y = mean,
           colour = parameter, fill = parameter)) +
  geom_line() +
  geom_hline(data = mle.estimates,
             aes(yintercept = value, linetype = MLE)) +
  scale_linetype_manual(values = c("dotted", "dashed")) + 
  geom_ribbon(aes(ymin = lwr, ymax = upr), alpha = .2) +
  scale_color_colorblind() +
  scale_fill_colorblind() +
  scale_y_continuous("", expand = c(0, 0)) +
  scale_x_continuous(expression(w), expand = c(0, 0)) +
  facet_wrap(parameter~., scales = "free_y") +
  theme_bw(base_size = 16)
```
As we can see, the actual posterior estimates do not seem to be very sensitive to
the value of $w \in [0.1, 0.9]$.
This might prompt the curious to ask whether this would change for small values
of $w$.
Let's investigate:
```{r smallRMAP, cache=TRUE}

ws.small <- c(10^{-(6:2)}, 2:9/100)

system.time(
  all.fits.small <- lapply(ws.small, rmap_fit_and_report)
)

all.summaries.small <- lapply(1:length(all.fits.small),
                              function(i) {
                                out <- tibble(get_summaries(fit = all.fits.small[[i]], base.pars),
                                              name = paste0("w=", ws.small[i]))
                                return(out)
                              })

results.small <- do.call(rbind, all.summaries.small) 
results.small$w <- as.numeric(gsub("w=", "", results.small$name))
results.small <- results.small %>% rename(lwr = q5)
results.small <- results.small %>% rename(upr = q95)
results.small <- results.small %>% rename(parameter = variable)

```
Again, let's plot 
```{r RMAPplot_small, echo=FALSE, cache=TRUE,  fig.align='left', fig.height=6, fig.width=9, fig.fullwidth=TRUE}

ggplot(data = results.small,
       aes(x = w, y = mean,
           colour = parameter, fill = parameter)) +
  geom_line() +
  geom_hline(data = mle.estimates,
             aes(yintercept = value, linetype = MLE)) +
  scale_linetype_manual(values = c("dotted", "dashed")) + 
  geom_ribbon(aes(ymin = lwr, ymax = upr), alpha = .2) +
  scale_color_colorblind() +
  scale_fill_colorblind() +
  scale_y_continuous("", expand = c(0, 0)) +
  scale_x_continuous(expression(w),
                     trans = log_trans(),
                     breaks = scales::trans_breaks("log10", function(x) 10^x),
                     labels = scales::trans_format("log10", scales::math_format(10^.x))) +
  facet_wrap(parameter~., scales = "free_y") +
  theme_bw(base_size = 16)
```
and voilà!
Now we do see that the posterior resulting from the RMAP is indeed sensitive
to $w$, as long as that value is well below $0.1$.

# Power prior (PP)

We now turn our attention to the power prior (PP, @Ibrahim2000), which raises 
the likelihood of the historical data to a discounting factor, $a_0$:
\begin{equation}
\pi_{\text{PP}}(\theta) \propto L(D_0 \mid \theta)^{a_0}\pi_0(\theta).
\end{equation}

```{r ppfits, cache=TRUE}

pp_fit_and_report <- function(a){
  cat("Doing a0=", a, "\n")
  fit.pp <- glm.pp(
    formula,
    family,
    data.list = the.data,
    a0 = a,
    parallel_chains = 4,
    iter_warmup = warmup,
    iter_sampling = samples,
    show_messages = FALSE,
    show_exceptions = FALSE,
    refresh = 0
  )
  return(fit.pp)
}

as <- sort(c(ws.small, ws.regular))  

system.time(
  all.pp.fits <- lapply(as, pp_fit_and_report)
)

all.summaries.pp <- lapply(1:length(all.pp.fits),  function(i) {
  out <- tibble(get_summaries(fit = all.pp.fits[[i]], base.pars),
                name = paste0("a_0=", as[i]))
  return(out)
})

results.pp <- do.call(rbind, all.summaries.pp) 
results.pp$a0 <- as.numeric(gsub("a_0=", "", results.pp$name))
results.pp <- results.pp %>% rename(lwr = q5)
results.pp <- results.pp %>% rename(upr = q95)
results.pp <- results.pp %>% rename(parameter = variable)
```
We can now plot the results:
```{r PPplot_complete, echo=FALSE, cache=TRUE,  fig.align='left', fig.height=6, fig.width=9, fig.fullwidth=TRUE}
ggplot(data = results.pp,
       aes(x = a0, y = mean,
           colour = parameter, fill = parameter)) +
  geom_line() +
  geom_hline(data = mle.estimates,
             aes(yintercept = value, linetype = MLE)) +
  scale_linetype_manual(values = c("dotted", "dashed")) + 
  geom_ribbon(aes(ymin = lwr, ymax = upr), alpha = .2) +
  scale_color_colorblind() +
  scale_fill_colorblind() +
  scale_y_continuous("", expand = c(0, 0)) +
  scale_x_continuous(expression(a[0]),
                     trans = log_trans(),
                     breaks = scales::trans_breaks("log10", function(x) 10^x),
                     labels = scales::trans_format("log10", scales::math_format(10^.x))) +
  facet_wrap(parameter~., scales = "free_y") +
  theme_bw(base_size = 16)
```
only to realise that the sensitivity is in the $(0.1, 1)$ interval.
Let's slightly re-do the plot:

```{r PPplot_small, echo=FALSE, cache=TRUE,  fig.align='left', fig.height=6, fig.width=9, fig.fullwidth=TRUE}
ggplot(data = results.pp,
       aes(x = a0, y = mean,
           colour = parameter, fill = parameter)) +
  geom_line() +
  geom_hline(data = mle.estimates,
             aes(yintercept = value, linetype = MLE)) +
  scale_linetype_manual(values = c("dotted", "dashed")) + 
  geom_ribbon(aes(ymin = lwr, ymax = upr), alpha = .2) +
  scale_color_colorblind() +
  scale_fill_colorblind() +
  scale_y_continuous("", expand = c(0, 0)) +
  scale_x_continuous(expression(a[0]),
                     expand = c(0, 0)) +
  facet_wrap(parameter~., scales = "free_y") +
  theme_bw(base_size = 16)
```

Now we can clearly see that the posteriors seem to start converging when $a_0 > 0.3$,
although the picture is not as clear cut as for the RMAP.

Now, let's take a look at prior sensitivity in normalised versions of the power prior, the NPP and NAPP.

First, let's elicit Beta priors with a given prior mean and maximum entropy (subject to having the right mean)
```{r}
# remotes::install_github("")
elicit_beta_mean_maxent <- 
function (m0) 
{
  # taken from maxbiostat/logPoolR
    eta <- (1/m0 - 1)
    Q <- function(a) {
        b <- a * eta
        out <- entropy_beta(a, b)
        return(-out)
    }
    Opt <- optimize(Q, lower = 0, upper = 10000)
    a <- Opt$minimum
    b <- a * eta
    if (a < 0 || b < 0) {
        warning("Warning: at least one of the obtained parameters is not valid")
    }
    return(list(a = a, b = b))
}
hyperpars <- lapply(as, elicit_beta_mean_maxent)
K <- length(hyperpars)
prior.quantiles <- vector(K, mode = "list")
Alpha <- .95
for(k in 1:K) {
  prior.quantiles[[k]] <- tibble(median = qbeta(p = 1/2,
                                                shape1 = hyperpars[[k]]$a,
                                                shape2 = hyperpars[[k]]$b), 
                                 lwr = qbeta(p = (1 - Alpha)/2,
                                             shape1 = hyperpars[[k]]$a,
                                             shape2 = hyperpars[[k]]$b),
                                 upr = qbeta(p = (1 + Alpha)/2,
                                             shape1 = hyperpars[[k]]$a,
                                             shape2 = hyperpars[[k]]$b))
  prior.quantiles[[k]]$a0 <- as[k]
  
  prior.quantiles.df <- do.call(rbind, prior.quantiles)
}
```
Now let's plot these (hyper-) priors on $a_0$:

```{r priorPlotA0, echo=FALSE, cache=TRUE,  fig.align='left', fig.height=6, fig.width=9, fig.fullwidth=TRUE}
ggplot(data = prior.quantiles.df,
       aes(x = a0, y = median)) +
  geom_line() +
  geom_ribbon(aes(ymin = lwr, ymax = upr), alpha = .2) +
  scale_y_continuous(expression(a[0]), expand = c(0, 0)) +
  scale_x_continuous("Prior mean",
                     expand = c(0, 0)) +
  ggtitle("Maximum entropy Beta prior on the discounting parameter") +
  geom_abline(slope = 1, intercept = 0, linetype = "longdash") +
  theme_bw(base_size = 16)
```
These seem wide enough.
Now, let's prepare what we need to fit the normalised power prior (NPP) under each of these priors.


```{r normconst, cache = TRUE}
a0.lognc <- list()
a0.lognc.hdbayes <- data.frame(a0 = as)

glm_lp <- hdbayes:::glm_lp
get_lp2mean <- hdbayes:::get_lp2mean
bernoulli_glm_lp <- hdbayes:::bernoulli_glm_lp


## call created function
for (i in 2:length(the.data)) {
  histdata = the.data[[i]]
  ## wrapper to obtain log normalizing constant in parallel package
  logncfun <- function(a0, ...) {
    glm.npp.lognc(
      formula = formula,
      family = family,
      a0 = a0,
      histdata = histdata,
      show_messages = FALSE,
      show_exceptions = FALSE,
      ...
    )
  }
  cl <- parallel::makeCluster(10)
  parallel::clusterSetRNGStream(cl, 123)
  parallel::clusterExport(
    cl,
    varlist = c(
      'formula',
      'family',
      'histdata',
      'glm.npp.lognc',
      'glm_lp',
      'get_lp2mean',
      'bernoulli_glm_lp'
    )
  )
  a0.lognc[[i - 1]] <- parallel::parLapply(
    cl = cl,
    X = as,
    fun = logncfun,
    iter_warmup = warmup,
    iter_sampling = samples,
    chains = 1,
    refresh = 0
  )
  
  stopCluster(cl)
  a0.lognc[[i - 1]] <- data.frame(do.call(rbind, a0.lognc[[i - 1]]))
  a0.lognc.hdbayes <- cbind(a0.lognc.hdbayes, a0.lognc[[i - 1]]$lognc)
  colnames(a0.lognc.hdbayes)[i] <- paste0("lognc_hist", i - 1)
}

a0.lognc <- a0.lognc.hdbayes$a0
lognc    <- as.matrix(a0.lognc.hdbayes[, -1, drop = F])
```
Now we will actually fit the NPP under each of these priors.

```{r nppfits, cache=TRUE}

npp_fit_and_report <- function(k){
  cat("Doing a0=", as[k], "\n")
  fit.npp <- glm.npp(
    formula,
    family,
    data.list = the.data,
    a0.lognc = a0.lognc,
    lognc = lognc,
    a0.shape1 =  hyperpars[[k]]$a, a0.shape2 = hyperpars[[k]]$b,
    parallel_chains = 4,
    iter_warmup = warmup,
    iter_sampling = samples,
    show_messages = FALSE,
    show_exceptions = FALSE,
    refresh = 0
  )
  return(fit.npp)
}

system.time(
  all.npp.fits <- lapply(seq_along(as), npp_fit_and_report)
)

all.summaries.npp <- lapply(1:length(all.npp.fits),  function(i) {
  out <- tibble(get_summaries(fit = all.npp.fits[[i]], base.pars),
                name = paste0("a_0=", as[i]))
  return(out)
})

results.npp <- do.call(rbind, all.summaries.npp) 
results.npp$prior_mean_a0 <- as.numeric(gsub("a_0=", "", results.npp$name))
results.npp <- results.npp %>% rename(lwr = q5)
results.npp <- results.npp %>% rename(upr = q95)
results.npp <- results.npp %>% rename(parameter = variable)
```

```{r NPPplot_all, echo=FALSE, cache=TRUE,  fig.align='left', fig.height=6, fig.width=9, fig.fullwidth=TRUE}
ggplot(data = results.npp,
       aes(x = prior_mean_a0, y = mean,
           colour = parameter, fill = parameter)) +
  geom_line() +
  geom_hline(data = mle.estimates,
             aes(yintercept = value, linetype = MLE)) +
  scale_linetype_manual(values = c("dotted", "dashed")) + 
  geom_ribbon(aes(ymin = lwr, ymax = upr), alpha = .2) +
  scale_color_colorblind() +
  scale_fill_colorblind() +
  scale_y_continuous("", expand = c(0, 0)) +
  scale_x_continuous(expression("Prior mean for"~a[0]),
                     expand = c(0, 0)) +
  facet_wrap(parameter~., scales = "free_y") +
  theme_bw(base_size = 16)
```
And now we will repeat the exercise with the normalised asymptotic power prior (NAPP).

```{r nappfits, cache=TRUE}

napp_fit_and_report <- function(k){
  cat("Doing a0=", as[k], "\n")
  fit.napp <- glm.napp(
    formula,
    family,
    data.list = the.data,
    a0.shape1 =  hyperpars[[k]]$a, a0.shape2 = hyperpars[[k]]$b,
    parallel_chains = 4,
    show_messages = FALSE,
    show_exceptions = FALSE,
    iter_warmup = warmup,
    iter_sampling = samples,
    refresh = 0
  )
  return(fit.napp)
}

system.time(
  all.napp.fits <- lapply(seq_along(as), napp_fit_and_report)
)

all.summaries.napp <- lapply(1:length(all.napp.fits),  function(i) {
  out <- tibble(get_summaries(fit = all.napp.fits[[i]], base.pars),
                name = paste0("a_0=", as[i]))
  return(out)
})

results.napp <- do.call(rbind, all.summaries.napp) 
results.napp$prior_mean_a0 <- as.numeric(gsub("a_0=", "", results.napp$name))
results.napp <- results.napp %>% rename(lwr = q5)
results.napp <- results.napp %>% rename(upr = q95)
results.napp <- results.napp %>% rename(parameter = variable)
```

```{r NAPPplot_all, echo=FALSE, cache=TRUE,  fig.align='left', fig.height=6, fig.width=9, fig.fullwidth=TRUE}
ggplot(data = results.napp,
       aes(x = prior_mean_a0, y = mean,
           colour = parameter, fill = parameter)) +
  geom_line() +
  geom_hline(data = mle.estimates,
             aes(yintercept = value, linetype = MLE)) +
  scale_linetype_manual(values = c("dotted", "dashed")) + 
  geom_ribbon(aes(ymin = lwr, ymax = upr), alpha = .2) +
  scale_color_colorblind() +
  scale_fill_colorblind() +
  scale_y_continuous("", expand = c(0, 0)) +
  scale_x_continuous(expression("Prior mean for"~a[0]),
                     expand = c(0, 0)) +
  facet_wrap(parameter~., scales = "free_y") +
  theme_bw(base_size = 16)
```

Finally, we will look at the posterior distribution for $a_0$:

```{r posteriorPlotA0, echo=FALSE, cache=TRUE,  fig.align='left', fig.height=6, fig.width=9, fig.fullwidth=TRUE}

npp.a0.posterior <- do.call(rbind,
                            lapply(seq_along(all.npp.fits), function(k){
                              tibble(median = mean(all.npp.fits[[k]]$a0_hist_1),
                                     lwr = quantile(all.npp.fits[[k]]$a0_hist_1, probs = (1 - Alpha)/2),
                                     upr = quantile(all.npp.fits[[k]]$a0_hist_1, probs = (1 + Alpha)/2),
                                     a0 = as[k],
                                     model = "NPP")
                            }))


napp.a0.posterior <- do.call(rbind,
                             lapply(seq_along(all.napp.fits), function(k){
                               tibble(median = median(all.napp.fits[[k]]$a0_hist_1),
                                      lwr = quantile(all.napp.fits[[k]]$a0_hist_1, probs = (1 - Alpha)/2),
                                      upr = quantile(all.napp.fits[[k]]$a0_hist_1, probs = (1 + Alpha)/2),
                                      a0 = as[k],
                                      model = "NAPP")
                             }))

prior.quantiles.df$model <- "prior"

a0.posterior.df <- do.call(rbind, list(prior.quantiles.df, npp.a0.posterior, napp.a0.posterior))

ggplot(data = a0.posterior.df,
       aes(x = a0, y = median, colour = model, fill = model)) +
  geom_line() +
  geom_ribbon(aes(ymin = lwr, ymax = upr), alpha = .2) +
  scale_y_continuous(expression(a[0]), expand = c(0, 0)) +
  scale_x_continuous("Prior mean",
                     expand = c(0, 0)) +
  ggtitle("Posterior quantiles for the discounting parameter") +
  geom_abline(slope = 1, intercept = 0, linetype = "longdash") +
  theme_bw(base_size = 16)
```
from which we conclude that the posterior for $a_0$ does not change much relative to its prior.

# References {-}

<div id="refs"></div>

# Computing environment

```{r env}
sessionInfo()
```

